---
title: "Creating a species composition matrix for further analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a species composition matrix for further analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DivInsight)
data("Colombia")
library(vegan)
```

A species composition matrix can be created using the data from a specific location. This is useful for further analyses.

```{r}
# use coordinates to clusterise data 
Colombia_coordinate_ref <- clusterise_sites(
  
  dataframe = 
    subset_by_coordinate_ref(
      dataframe = Colombia,
      coordinate_reference = c(-73.487520, 7.539986),
      distance_threshold = 50000
    ),
  
  cluster_min_length = 30 
  
)
```

`site_regroup()` can be used to group sites by a specified radius in metres.

This method is a lot faster than running `clusterise_sites()` again with new parameters.

```{r}
# regroup sites with a radius of 1km
cref_1km <- site_regroup(
  
  clusterised_object = Colombia_coordinate_ref,
  regroup_radius = 1000
  
)
```

Species tables can be produced for each site using `generate_spec_tables()`. The number of individuals and species in each tables can be specified.

Each table will consist of one row pertaining to the site's group and each column will be a species name.

```{r}
cref_1km_spectables <- generate_spec_tables(
  
  clusterised_object = cref_1km,
  min_individuals = 30,
  min_species = 10
  
)
```

The name of each table can be viewed where the first number represents which group the site is in.

The following information is a date. Users should be aware of vastly different times between species tables as well as seasonal differences that may hinder the results of their analysis.

using the $ operator is also a useful way to select which species tables to include in their species composition matrix.


```{r}
# view the names of each table
print(names(cref_1km_spectables))

# store the chosen species tables into a single list 
species_table_list <- list(
  
  cref_1km_spectables$`17.2022-04-04`,
  cref_1km_spectables$`29.2022-04-02`,
  cref_1km_spectables$`16.2022-04-01`,
  cref_1km_spectables$`26.2022-03-31`,
  cref_1km_spectables$`1.2022-03-30`,
  
  cref_1km_spectables$`28.2022-03-29`,
  cref_1km_spectables$`18.2022-03-28`,
  cref_1km_spectables$`27.2022-03-26`,
  cref_1km_spectables$`9.2022-03-25`,
  cref_1km_spectables$`25.2022-03-23`,
  
  cref_1km_spectables$`8.2022-03-21`,
  cref_1km_spectables$`7.2022-03-19`,
  cref_1km_spectables$`24.2022-03-07`,
  cref_1km_spectables$`23.2022-03-06`,
  cref_1km_spectables$`19.2022-03-05`,
  
  cref_1km_spectables$`22.2022-02-27`,
  cref_1km_spectables$`21.2022-02-15`,
  cref_1km_spectables$`20.2022-02-14`
  
)     

# use the generate_speccomm to generate a species composition matrix
SCM1 <- generate_speccomm(species_table_list)
```


The choice to change the name of each row is optional.

```{r}
# change the row names of the matrix
SCM1 <- as.data.frame(SCM1)

row.names(SCM1) <- c(
  
  "17.2022-04-04",
  "29.2022-04-02",
  "16.2022-04-01",
  "26.2022-03-31",
  "1.2022-03-30",
  
  "28.2022-03-29",
  "18.2022-03-28",
  "27.2022-03-26",
  "9.2022-03-25",
  "25.2022-03-23",
  
  "8.2022-03-21",
  "7.2022-03-19",
  "24.2022-03-07",
  "23.2022-03-06",
  "19.2022-03-05",
  
  "22.2022-02-27",
  "21.2022-02-15",
  "20.2022-02-14"

)

SCM1 <- as.matrix(SCM1)

```


There are many analyses that can be done with a species composition matrix but here a species accumulation curve using the vegan package will be demonstrated.

```{r}
# create a species accumulation curve 
speccurve1 <- specaccum(SCM1, method = "random")

# plot the species accumulation curve
plot(speccurve1, ci.type="poly", 
     col="blue", 
     ci.col="lightblue",
     main = "Species accumulation curve for 15 Sites in coordinate reference site #1"
)

# view the predictions from the species accumulation curve
print(speccurve1)
```
